Finding and controling biases in dark matter halo concentrations estimates

### 2016-07-06 paper/ms.tex
reformulando conclusiones

We notice first that the results from the density method have a
systematic $15\%$offset from the velocity methods.  
This offset was already presented by \citep{Prada2012} for low
concentrations ($c<6$) and high ($M_h>10^{12}\hMsun$) halo masses.  
Recently \citep{Klypin2016} summarized results for the
mass-concentration relationship coming from different methods and
datasets to show that similar systematic offsets are present.
However, to the best of our knowledge, a systematic and modern study of
the mass-concentration relationship spanning a wide range in mass,
using a fixed dataset but different fitting methods has not been
presented in the literature.

@@

In this paper we quantified biases on concentration numerical estimates by
bootstrapping dark matter halos resolved with a  large number of
particles ($\sim 10^5$).
We found that methods commonly used in the literature present a large
bias in the concentration by overestimating it by factors of
$15\%$-$20\%$ for halos with $200$ particles, or $6\%$-$8\%$ for halos
with $500$ particles. 
This procedure is a new way to quantify the bias in concentration
estimates.

The scaling of the bias as a function of particle number was found using
two very different kinds of simulations, one cosmological and the
other of an individual high resolution halo.

This results motivated the introduction of a new method to estimate the
concentration with a lower bias at low particle numbers.
The method we presented is based on the integrated mass profile. 
To show its performance on simulation data Ww also implemented a
fitting procedure based on a Markov Chain Monte Carlo algorithm.
In the bootstrapping experiment this method showed a bias on the order
of $2\%-3\%$ for halos with $200$ particles and less than $1\%$ for
halos with $500$ particles or more.

We also used the N-body simulation to estimate the impact on the mass
concentration relationship of the new method to estimate
concentration.   
Although the three methods are in broad agreement within their
uncertainties there are some noticeable differences.  
The first difference is that the density method produces
systematically higher concentrations by a factor of $15\%$ compared to
the velocity method.
This systematic offset has been reported in different datasets and
techniques to measure concentrations, without any conclusive
explanation for its origin \citep{Klypin2016}.
The second difference is that the velocity and integrated mass methods
start to differ for masses below $10^{12}\hMsun$.  
We found that a correction of the results by the bias factor found
using the same N-body data can bring these two methods into
agreement.

These are encouraging results.  
They show that using the integrated mass profile to find the Dark
Matter halo concentration is a tool deserving deeper scrutiny.  
Further tests with larger simulated volumes, varying numerical
resolution and different density profiles are the next natural step to
explore the full potential of this new method.

@@

\bibliographystyle{apj}
\bibliography{references}

>>>>

We notice first that the results from the density method have a
systematic $15\%$offset from the velocity methods.  
This offset was already presented by \citep{Prada2012} for low
concentrations ($c<6$) and high ($M_h>10^{12}\hMsun$) halo masses.  
Recently \citep{Klypin2016} summarized results for the
mass-concentration relationship coming from different methods and
datasets to show that similar systematic offsets are present.
\citep{2014MNRAS.441.3359D} studied the mass-concentration
relationship using the maximum velocity and density methods and did
not report any significant difference. However, they implemented a
modified version of the velocity algorithm that uses the data binning
obtained for the density algorithm.

@@

In this paper we quantified biases on concentration numerical estimates by
bootstrapping dark matter halos resolved with a  large number of
particles ($\sim 10^5$).
We found that methods commonly used in the literature present a large
bias by overestimating concentrations by factors of $13\%$-$20\%$ for
halos with $200$ particles, or $7\%$-$10\%$ for halos with $500$
particles.  
This procedure is a new way to quantify the bias in concentration
estimates with the advantage that it works without having to run new
simulations.

These results motivated us to introduction a method with a robust
performace at low particle numbers.
This method is based on the integrated mass profile. 
The results of the algorithm on the same data sample showed a bias  of
$2\%-3\%$ for halos with $200$ particles and less than $1\%$ for halos
with $500$ particles or more.

We also measured the mass concentration relationship. 
The three methods are in broad agreement within the statistical
uncertainties, although there are some noticeable differences.
The first difference is that the density method produces
systematically higher concentrations by a factor of $15\%$ compared to
the velocity method.
This systematic offset has been reported before with the same dataset 
\citep{Prada2012} and with different simulations \citep{Klypin2016}
without any conclusive explanation for its origin. 
The second difference is that the velocity and integrated mass methods
start to differ for masses below $10^{12}\hMsun$.  
We found that correcting the mean concentration by the mean bias
factor brings these two techniques into agreement

These results show that using the integrated mass profile to find the
Dark Matter halo concentration is a tool deserving deeper scrutiny.   
Further tests with larger simulated volumes, varying numerical
resolution and different density profiles are the next natural step to
explore the full potential of this new method.

@@

\bibliographystyle{apj}
%\bibliography{references}
\begin{thebibliography}{15}
\expandafter\ifx\csname natexlab\endcsname\relax\def\natexlab#1{#1}\fi

\bibitem[{{Diemand} {et~al.}(2008){Diemand}, {Kuhlen}, {Madau}, {Zemp},
  {Moore}, {Potter}, \& {Stadel}}]{2008Natur.454..735D}
{Diemand}, J., {Kuhlen}, M., {Madau}, P., {Zemp}, M., {Moore}, B., {Potter},
  D., \& {Stadel}, J. 2008, \nat, 454, 735

\bibitem[{{Duffy} {et~al.}(2008){Duffy}, {Schaye}, {Kay}, \& {Dalla
  Vecchia}}]{Duffy2008}
{Duffy}, A.~R., {Schaye}, J., {Kay}, S.~T., \& {Dalla Vecchia}, C. 2008,
  \mnras, 390, L64

\bibitem[{{Dutton} \& {Macci{\`o}}(2014)}]{2014MNRAS.441.3359D}
{Dutton}, A.~A., \& {Macci{\`o}}, A.~V. 2014, \mnras, 441, 3359

\bibitem[{{Foreman-Mackey} {et~al.}(2013){Foreman-Mackey}, {Hogg}, {Lang}, \&
  {Goodman}}]{emcee}
{Foreman-Mackey}, D., {Hogg}, D.~W., {Lang}, D., \& {Goodman}, J. 2013, \pasp,
  125, 306

\bibitem[{{Klypin} {et~al.}(2016){Klypin}, {Yepes}, {Gottl{\"o}ber}, {Prada},
  \& {He{\ss}}}]{Klypin2016}
{Klypin}, A., {Yepes}, G., {Gottl{\"o}ber}, S., {Prada}, F., \& {He{\ss}}, S.
  2016, \mnras, 457, 4340

\bibitem[{{Ludlow} {et~al.}(2014){Ludlow}, {Navarro}, {Angulo},
  {Boylan-Kolchin}, {Springel}, {Frenk}, \& {White}}]{Ludlow2014}
{Ludlow}, A.~D., {Navarro}, J.~F., {Angulo}, R.~E., {Boylan-Kolchin}, M.,
  {Springel}, V., {Frenk}, C., \& {White}, S.~D.~M. 2014, \mnras, 441, 378

\bibitem[{{Macci{\`o}} {et~al.}(2008){Macci{\`o}}, {Dutton}, \& {van den
  Bosch}}]{Maccio2008}
{Macci{\`o}}, A.~V., {Dutton}, A.~A., \& {van den Bosch}, F.~C. 2008, \mnras,
  391, 1940

\bibitem[{{More} {et~al.}(2011){More}, {Kravtsov}, {Dalal}, \&
  {Gottl{\"o}ber}}]{More2011}
{More}, S., {Kravtsov}, A.~V., {Dalal}, N., \& {Gottl{\"o}ber}, S. 2011, \apjs,
  195, 4

\bibitem[{{Mu{\~n}oz-Cuartas} {et~al.}(2011){Mu{\~n}oz-Cuartas}, {Macci{\`o}},
  {Gottl{\"o}ber}, \& {Dutton}}]{Munoz2011}
{Mu{\~n}oz-Cuartas}, J.~C., {Macci{\`o}}, A.~V., {Gottl{\"o}ber}, S., \&
  {Dutton}, A.~A. 2011, \mnras, 411, 584

\bibitem[{{Navarro} {et~al.}(1997){Navarro}, {Frenk}, \& {White}}]{NFW}
{Navarro}, J.~F., {Frenk}, C.~S., \& {White}, S.~D.~M. 1997, \apj, 490, 493

\bibitem[{{Navarro} {et~al.}(2010){Navarro}, {Ludlow}, {Springel}, {Wang},
  {Vogelsberger}, {White}, {Jenkins}, {Frenk}, \& {Helmi}}]{Navarro2010}
{Navarro}, J.~F., {Ludlow}, A., {Springel}, V., {Wang}, J., {Vogelsberger}, M.,
  {White}, S.~D.~M., {Jenkins}, A., {Frenk}, C.~S., \& {Helmi}, A. 2010,
  \mnras, 402, 21

\bibitem[{{Neto} {et~al.}(2007){Neto}, {Gao}, {Bett}, {Cole}, {Navarro},
  {Frenk}, {White}, {Springel}, \& {Jenkins}}]{Neto2007}
{Neto}, A.~F., {Gao}, L., {Bett}, P., {Cole}, S., {Navarro}, J.~F., {Frenk},
  C.~S., {White}, S.~D.~M., {Springel}, V., \& {Jenkins}, A. 2007, \mnras, 381,
  1450

\bibitem[{{Prada} {et~al.}(2012){Prada}, {Klypin}, {Cuesta}, {Betancort-Rijo},
  \& {Primack}}]{Prada2012}
{Prada}, F., {Klypin}, A.~A., {Cuesta}, A.~J., {Betancort-Rijo}, J.~E., \&
  {Primack}, J. 2012, \mnras, 423, 3018

\bibitem[{{Riebe} {et~al.}(2013){Riebe}, {Partl}, {Enke}, {Forero-Romero},
  {Gottl{\"o}ber}, {Klypin}, {Lemson}, {Prada}, {Primack}, {Steinmetz}, \&
  {Turchaninov}}]{2013AN....334..691R}
{Riebe}, K., {Partl}, A.~M., {Enke}, H., {Forero-Romero}, J., {Gottl{\"o}ber},
  S., {Klypin}, A., {Lemson}, G., {Prada}, F., {Primack}, J.~R., {Steinmetz},
  M., \& {Turchaninov}, V. 2013, Astronomische Nachrichten, 334, 691

\bibitem[{{Stadel}(2001)}]{2001PhDT........21S}
{Stadel}, J.~G. 2001, PhD thesis, UNIVERSITY OF WASHINGTON

\end{thebibliography}
